UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-05-29 ms/main.tex
minor text tweaks

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{\textbf{Four methods of dimension reduction of 1KGP genotype data with population labels.} (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary. Since the algorithms prioritize local distances, long distances between clusters are not meaningful.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

@@

The HRS forms several large groupings and clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}). The clarity of the interconnected clusters varies by parameterization, but they consistently form a large, roughly connected group. \sgcomment{Last sentence is a bit vague. Remove?}

To investigate possible ancestries related to populations in the 1KGP we took two approaches. In the first we generated PC axes and a UMAP embedding using HRS and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}). In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}). Both approaches reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population. One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

@@

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.jpeg}%
\caption{}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{\textbf{The UKBB coloured by self-reported ethnic background.} \textbf{(a)} The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. \textbf{(b)} UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

The largest body in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population concentrates in one part of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group". Using data on countries of birth, we identified many finer groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines have coordinates in the south east of the UK, where the city of London isa located, likely reflecting its high migrant population. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB. %The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.

>>>>

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{\textbf{Four methods of dimension reduction of 1KGP genotype data with population labels.} (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary. Since the algorithms prioritize local distances, long distances between clusters are not meaningful.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

@@

The HRS forms several large clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}). Projecting 1KGP populations onto the embedding reveals substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}). One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

%To investigate possible ancestries related to populations in the 1KGP we took two approaches. In the first we generated PC axes and a UMAP embedding using HRS and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}). In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}). Both approaches reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population. One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

@@

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.jpeg}%
\caption{}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{\textbf{The UKBB coloured by self-reported ethnic background.} \textbf{(a)} The first two principal components, showing the usual triangle with vertices corresponding to African, Asian, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. \textbf{(b)} UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

The largest cluster in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population forms a sub-cluster, but many individuals are also scattered throughout the British-identifying population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian population; Indian, Pakistani, and Bangladeshi populations form a closely bound cluster as well. The East Asian and South Asian populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group". Using data on countries of birth, we identified many finer groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino clusters being projected near Chinese clusters.

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines have coordinates in the south east of the UK, where the city of London is located, likely reflecting its high migrant population. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB. %The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.
